Quantifying multi-modal competition using spatial availability

### 2023-07-30 NECTAR-Coimbra-Presentation.Rmd
Presentation file commits pt.1.

---
title: "Multi-modal spatial availability: a research proposal"
subtitle: "CIT - Tenerife 2023"
author: "Anastasia Soukhov, J. Tarriño Ortiz, J. A. Soria-Lara, A. Páez"
date: "2022-11-10"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      slideNumberFormat: "%current% / 14" 
      highlightStyle: github
      highlightLines: true
      ratio: '16:9'
      countIncrementalSlides: false
      beforeInit: "macros.js"
---
```{r xaringan-themer, include=FALSE, warning=FALSE}
#this themer package allows us to change the .css themes through this chunk. Easier than changing the .css directly.
library(xaringanthemer)
style_mono_accent(
  base_color = "#7A003C", #maroon
  link_color = "#5E6A71", #grey
  text_bold_color = "#FDBF57", #gold
  text_slide_number_color = "#FDBF57", #gold
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "300", "300i"),
  code_font_google   = google_font("Fira Mono")
)
```

```{r load-packages, include=FALSE, cache=FALSE}
library(dplyr)
library(sf)
library(ggplot2)
library(tidyverse)
library(units)
library(patchwork)
library(ggspatial)
library(ggforce)
library(geomtextpath)
library(knitr)
library(kableExtra)
options(scipen = 999)
```

@@

#write the dataframe long.
od_tt_m <- data.frame(i = c("A", "A", "A", "A", "A", "A",
                            "B", "B", "B", "B", "B", "B",
                            "C", "C", "C", "C", "C", "C"), # Three origins
                    j = c("1", "2", "3", "1", "2", "3", 
                          "1", "2", "3", "1", "2", "3", 
                          "1", "2", "3", "1", "2", "3"), # Three destinations
                    m = c("pt", "pt", "pt", "car", "car", "car",
                          "pt", "pt", "pt", "car", "car", "car",
                          "pt", "pt", "pt", "car", "car", "car"),
                    tt = c(15, 30, 100, 10, 25, 80, #travel time, z-mode (original) AND THEN x-mode (the faster mode)
                           30, 15, 100, 25,  10,  80,
                           100, 100, 15, 80, 80, 10), 
                    pop = c(20000, 20000, 20000, 30000, 30000, 30000,
                            50000, 50000,50000, 100000, 100000, 100000,
                            3000, 3000, 3000, 7000,7000, 7000),
                    opp = c(100000, 100000, 10000, 100000, 100000, 10000,
                            100000, 100000, 10000, 100000, 100000, 10000,
                            100000, 100000, 10000, 100000, 100000, 10000)) 
```

@@

od_table <- data.frame(Center = c("A", "B", "C", "1", "2", "3"),
  Size = c("50,000 pop (40% pt)", "150,000 pop (33% pt)", "10,000 pop (30% pt)", 
           "100,000 jobs", "100,000 jobs", "10,000 jobs"))

@@

```{r, create-figure-with-toy-example, echo=FALSE, eval=FALSE}
toy_example_plot <- ggplot() + 
  # Plot centers
  geom_circle(data = centers,
             aes(x0 = x, 
                 y0 = y,
                 r = radius),
             color = "grey40") +
  annotate(geom = "label", 
           x=c(0.5, 0.6, 7),
           y=c(9, 1, 5.2), 
           label = c("Suburban", "Urban", "Satellite"), 
           size=4, 
           label.size=NA) +
  # Connect origins and destinations with curves
  geom_curve(data = od_lines,
               aes(x = x,
                   y = y,
                   xend = xend,
                   yend = yend,
                   linetype = group,
                   group = group),
             curvature = 0.20,
             linewidth = 1,
             color = "black") +
  # Text on curves
  geom_textcurve(data = od_lines,
               aes(x = x,
                   y = y,
                   xend = xend,
                   yend = yend,
                   linetype = group,
                   group = group,
                   label = time),
               size = 3.5,
             curvature = 0.20,
             color = "black") +
  # Plot origins and destinations
  geom_point(data = od,
            aes(size = size,
                x = x, 
                y = y, 
                color = type,
                shape = type)) +
  scale_size(range = c(4, 10)) +
  # Label origins and destinations
  geom_text(data = od, 
            aes(x, 
                y, 
                label=id), 
            size=4) +
  # Create table
  annotate(geom = "table",
           x = 9, 
           y = 9, 
           label = list(od_table),
           fill = "white",
           size=3) +
  # Theme the plot
  coord_fixed() + 
  theme_void() + guides(shape = "none", size = "none", color = "none", linetype ="none")

toy_example_plot
ggsave("Presentations/toy-example.png",dpi=600)
```

## Overview

**1.** Accessibility measures and multi-modal applications

**2.** Explaining **spatial availability** within multi-modal analysis

**3.** Policy scenario planning application: low emission zones in Madrid

**4.** Implications: the changes in spatial availability of opportunities, active mobility, GHG emissions

???
XXX
---
## Multi-modal accessibility measures

@@

#### Gravity-based accessibility (Hansen-type) - no competition

$$A_i^m = \sum_{j}O_jf(c_{ij})^m$$
-   $i$ is a set of origin locations
-   $j$ is a set of destination locations
-   $O_j$ is the number of opportunities at location $j$
-   $f(\cdot)$ is an impedance function of $c_{ij}$ (cost of moving between $i$ and $j$) for each mode $m$; if binary, the measure if _cumulative opportunities_

**E.g.**, used in _(Tahmasbi et al., 2019)_ to calculate $A_i^m$ (per mode). Variations on how it is combined across modes.

Suffers from the same interpretability issues _(Handy and Niemeier, 1997; Miller et al., 2018)_.

Does not include competition (demand:supply, mass effect) _(Shen, 1998; Merlin and Hu, 2018)_.

???
Accessibility:

- A way to relatively quantify the access that a spatial unit: a product of land-use and the population's means of transportation

- In other words, it measures the amount of _potential interaction_ a population has with opportunities _(Hansen, 1959)_

- It connects land-use and the population's means of transportation: the population are the people that live within the region and the opportunities are the destinations of interest to the population or of the transportation network itself.

-**Hansen**-type accessibility - the sum of opportunities multiplied by a function which measures of the cost of moving between i and j.

- this measure suffers from interpretability issues - what is travel-cost adjusted opportunities?

- it also doesn't consider competition, which has been shown to reflect empirical accessibility to competitive-opportunities worse than competitive accessibility measures.

Interpretability issues are not resolved when considering multiple modes.

- Used in _(Tahmasbi et al., 2019)_ to quantify access $A_i$ for $m$ walking, public transit, and private car modes to retail store destinations (different $f^m(c_{ij}^m)$ for each $m$). Variations on how it is combined across modes (if at all) for general accessibility. It varies so widely because interpretation is inconsistent.
- Suffers from the same interpretability issues _(Handy and Niemeier, 1997; Miller et al., 2018)_
- Does not include competition (demand:supply or the mass effect) _(Shen, 1998; Merlin and Hu, 2018)_

.pull-left[
#### Competitive accessibility (Shen-type) - no proportional allocation
Shen (1998) modified Hansen-type accessibility; popularized as the **two-step floating catchment area (2SFCA) method** by _(Luo and Wang, 2003)_

$$S_i^m = \sum_j \frac{O_jf(c_{ij})^m}{\sum_m D_j^m}$$

with:
$$D_j^m = \sum_i P_{i}^m f(c_{ij})^m$$
- $S_{i}^m$ is Shen-type accessibility for each mode $m$
- $D_j^m$ is the demand at $j$ from $P_{i}$ using each mode $m$

Generalized multi-model competition:
$$S_i = \sum_m \biggl(\frac{P_i^m}{P_i}\biggr) \sum_j \frac{O_jf(c_{ij})^m}{\sum_m \sum_i P_{i}^m f(c_{ij})^m}$$

E.g., used by _Tao et al., 2020_.

However, the resulting value is an accessibility _score_ so interpretation is obfuscated _(Soukhov et al., 2023)_.

???
- e.g., used by _Tao et al., 2020_ to calculate the $S_i^m$ for $m$ modes public transit and car to employment as well as the generalized accessibility $S_i$

- **Shen**-type accessibility - the sum of travel-cost adjusted opportunities (supply) divided by travel-cost adjusted demand! It considers travel-cost from both the opportunities perspective and the population side. It is done in 2 steps for the 2SFCA: "The first one computes the ratio between supply and demand of each health facility; the second step computes accessibility as the weighted sum of these ratios using a distance-decay function"

It should be mentioned that Weibull (1976) was the first to arrive to this same formula from axiomatic approach. Joseph and Bantock (1982) then applied it to health care access.

---
### Impedance functions - quantifying how subsets of people travel

.left-column[

$f(c_{ij})^m$ is a function of the cost of moving between $i$ and $j$ for modes $m$.

.right-column[]

---
count: false
### Impedance functions - quantifying how subsets of people travel

.left-column[

$f(c_{ij})^m$ is a function of the cost of moving between $i$ and $j$ for modes $m$.

$f(c_{ij})^{t.behav.1} = \exp(-0.1\cdot c_{ij})$

.right-column[
```{r comparison-impedance-functions-synthetic-example1, echo=FALSE, warning=FALSE, fig.width=8, fig.height=6}
data.frame(tt = seq(0, 100, 1)) %>%
  mutate(f1 = exp(-0.1 * tt)) %>%
  pivot_longer(cols = -tt,
               names_to = "Function",
               values_to = "f") %>%
  mutate(Function = case_when(Function == "f1" ~ "beta = 0.1")) %>%
  ggplot() +
  xlab(expression(c[ij]~"(travel time in minutes)")) +
  ylab(expression('f(c'[ij]*')')) +
  geom_line(aes(x = tt, 
                y = f,
                color = Function), size=1) +
  scale_color_manual(values = c("dark green"), label = c(expression("exp(-0.1 c"[ij]*")")))+
  scale_x_continuous(breaks = seq(0, 80, 20))+
  theme_minimal()+
  theme(axis.title = element_text(size=18),
        axis.text = element_text(size=16),
        legend.position = c(0.85, 0.9),
        legend.title = element_blank(),
        legend.text = element_text(size=18))
```
]

---
count: false
### Impedance functions - quantifying how subsets of people travel

.left-column[

$f(c_{ij})^m$ is a function of the cost of moving between $i$ and $j$ for modes $m$.

$f(c_{ij})^{\text{t.behav.1}} = \exp(-0.1\cdot c_{ij})$

$f(c_{ij})^{\text{t.behav.2}} = \exp(-0.6\cdot c_{ij})$

.right-column[
```{r comparison-impedance-functions-synthetic-example2, warning=FALSE, echo=FALSE, fig.width=8, fig.height=6}
data.frame(tt = seq(0, 100, 1)) %>%
  mutate(f1 = exp(-0.1 * tt),
         f2 = exp(-0.6 * tt)) %>%
  pivot_longer(cols = -tt,
               names_to = "Function",
               values_to = "f") %>%
  mutate(Function = case_when(Function == "f1" ~ "beta = 0.1",
                              Function == "f2" ~ "beta = 0.6")) %>%
  ggplot() +
  xlab(expression(c[ij]~"(travel time in minutes)")) +
  ylab(expression('f(c'[ij]*')')) +
  geom_line(aes(x = tt, 
                y = f,
                color = Function), size=1) +
  scale_color_manual(values = c("dark green", "#9999CC"), label = c(expression("exp(-0.1 c"[ij]*")"), expression("exp(-0.6 c"[ij]*")")))+
  scale_x_continuous(breaks = seq(0, 80, 20))+
  theme_minimal()+
  theme(axis.title = element_text(size=18),
        axis.text = element_text(size=16),
        legend.position = c(0.85, 0.9),
        legend.title = element_blank(),
        legend.text = element_text(size=18))
```
]

---
count: false
### Impedance functions - quantifying how subsets of people travel

.left-column[

$f(c_{ij})^m$ is a function of the cost of moving between $i$ and $j$ for modes $m$.

$f(c_{ij})^{\text{t.behav.1}} = \exp(-0.1\cdot c_{ij})$

$f(c_{ij})^{\text{t.behav.2}} = \exp(-0.6\cdot c_{ij})$

$f(c_{ij}) = 1$ if $c_{ij} \leq 30$ else 
  $f(c_{ij}) = 0$

.right-column[
```{r comparison-impedance-functions-synthetic-example3, warning=FALSE, echo=FALSE, fig.width=8, fig.height=6}
data.frame(tt = seq(0, 100, 1)) %>%
  mutate(f1 = exp(-0.1 * tt),
         f2 = exp(-0.6 * tt),
         f3 = ifelse(tt < 30, 1, 0)) %>%
  pivot_longer(cols = -tt,
               names_to = "Function",
               values_to = "f") %>%
  mutate(Function = case_when(Function == "f1" ~ "beta = 0.1",
                              Function == "f2" ~ "beta = 0.6",
                              Function == "f3" ~ "tt <= 30")) %>%
  ggplot() +
  xlab(expression(c[ij]~"(travel time in minutes)")) +
  ylab(expression('f(c'[ij]*')')) +
  geom_line(aes(x = tt, 
                y = f,
                color = Function), size=1) +
  scale_color_manual(values = c("dark green", "#9999CC", "#CC6666"), label = c(expression("exp(-0.1 c"[ij]*")"), expression("exp(-0.6 c"[ij]*")"), expression(c[ij]~"< 30 min")))+
  scale_x_continuous(breaks = seq(0, 80, 20))+
  theme_minimal()+
  theme(axis.title = element_text(size=18),
        axis.text = element_text(size=16),
        legend.position = c(0.85, 0.9),
        legend.title = element_blank(),
        legend.text = element_text(size=18))
```
]
---
### Simulated example - Hansen-type accessibility (no competition or proportional allocation consideration)

```{r, toy-example1, echo=FALSE}
knitr::include_graphics("Presentations/toy-example.png")
```

@@

$A_i^m = \sum_jO_jf(c_{ij})^m$

Impedance function: $f(c_{ij})^{\text{car}} = f(c_{ij})^{\text{pt}} =exp(-0.1 \cdot c_{ij})$;

@@

A <- od_tt_m |>
  group_by(i,m) |>
  summarize(A = sum(opp * f),
            .groups = "drop")

A %>% kable(format = "html",
        booktabs = TRUE,
        digits = 2)
```
]

???
- explain the map; blue and orange dots are origins and employment centers respectively. Travel time within each center is 15 mins, travel time between sub and urban is 30 mins and travel time to the satellite town from sub / urban is 100 mins.

@@

### Synthetic example - Shen-type accessibility (no proportional allocation consideration)

.pull-left[
```{r, echo=FALSE}
knitr::include_graphics("Presentations/toy-example.png")
```
]

@@

.pull-right[
$A_i^m$ and $S_i^m = \sum_j \frac{O_j f(c_{ij})^m}{\sum_m \sum_i P_{i}^m f(c_{ij})^m}$

@@

S_m <- od_tt_m %>%
  left_join(LOS_j |> 
              dplyr::select(j, LOS_j, opp, D_j),
            by = c("j")) %>%
  group_by(i,m) |>
  summarize(S_i_m = sum(LOS_j * f),
            pop = mean(pop),
            .groups = "drop") |>
  rename("S" = "S_i_m")

S_m <- A %>% 
  left_join(S_m, by=c("i","m"))

```
```{r, echo=FALSE, message=FALSE}
knitr::kable(S_m %>% dplyr::select(-pop),
             format = "html",
             booktabs = TRUE,
             digits = 2)
```
]

---
count: false
### Synthetic example - Shen-type accessibility (no proportional allocation consideration)

.pull-left[
```{r, echo=FALSE}
knitr::include_graphics("Presentations/toy-example.png")
```

@@

.pull-right[
$S_i = \sum_m \biggl(\frac{P_i^m}{P_i}\biggr) S_i^m$
```{r, echo=FALSE, message=FALSE}
S <- S_m %>%
  group_by(i) |>
  summarize(S = sum(pop*S)/sum(pop),
            .groups = "drop")

knitr::kable(S,
             format = "html",
             booktabs = TRUE,
             digits = 2)
```

@@

We can do better, let's use proportional allocation (NOTE: WHY??????????)

---
### Spatial availability - competitive _and_ proportional allocation

@@

- The proportional allocation factor combines two aspects of the problem, population and cost:

$F^{t,m}_{ij} = \frac{F^{pm}_{i} \cdot F^{cm}_{ij}}{\sum_{m=1}^M \sum_{i=1}^N F^{pm}_{i} \cdot F^{cm}_{ij}}$

- Where $F^{p,m}_{i}$ is population allocation factor:

$F^{p,m}_{i} = \frac{P^m_{i}}{\sum_{i} P^m_{i}}$

- And the $F^{c,m}_{ij}$ is a cost allocation factor:

$F^{c,m}_{ij} = \frac{f(c_{ij})^m}{\sum_{i} f(c_{ij})^m}$
]

@@

- Total spatial availability: $V_{i} = \sum_{m}^M V^m_{i}$
  - Considers **competition** and **proportionally allocates opportunities** (singly-constrained: $O = \sum_{j} O_j$)

- Total spatial availability per capita: $v_{i} = \sum_{m} \frac{V_i^m}{P_{i^m}}$
  - Equivalent to Shen-style accessibility

@@

  -supply of population and demand for opportunities are both travel-cost and population adjusted
  - why singly-constrained? the sum of all accessibility values for i equals all the opportunities for i to j (like in a singly constrained gravity model - it is attraction constrained)

---
### Spatial availability - competitive _and_ proportional allocation

.pull-left[
```{r, echo=FALSE}
knitr::include_graphics("Presentations/toy-example.png")
```
]

@@

$V^m_{i} = \sum_{j} O_j\ F^{t,m}_{ij}$ and $v^m_{i} = \frac{V^m_{i}}{P_{i}^m}$

@@

V_m <- od_tt_m %>%
  group_by(i,m) %>%
  summarize(V = sum(V_ij),
            v = V/first(pop))
```
```{r, echo=FALSE, message=FALSE}
knitr::kable(V_m,
             format = "html",
             booktabs = TRUE,
             digits = 2)
```

@@

---
count: false
### Spatial availability - competitive _and_ proportional allocation

.pull-left[
```{r, echo=FALSE}
knitr::include_graphics("Presentations/toy-example.png")
```
]

@@

.pull-right[
$S_i$ (eqv. to $v_i$) and $V_{i} = \sum_{m=1}^M V^m_{i}$

@@

V <- V %>% left_join(S, by=c("i")) |> transmute(S,V)

knitr::kable(V,
             format = "html",
             booktabs = TRUE,
             digits = 2)
```

@@

.pull-right[
```{r, echo=FALSE, out.width='50%', fig.align='center'}
knitr::include_graphics("Presentations/city-centre.png")
```

- November 2018 - LEZ Centro
- 2023 - LEZ within M-30
- 2024 - LEZ within Madrid Municipality 
]

???
A NEEED for multi-modal analysis methods that are interpretive and comparative -- proportional allocation achieves that!
---
### Multi-modal home-to-work accessibility in Madrid: 2018 travel survey

.pull-left[
```{r, echo=FALSE, out.width='120%',fig.align='center'}
knitr::include_graphics("Presentations/SA_i_V_zn1259_plot.png")
```
Spatial Availability $V_i$ at resolution Zona 1259
]

.pull-right[
```{r, echo=FALSE, out.width='120%', fig.align='center'}
knitr::include_graphics("Presentations/SA_i_V_zn208_plot.png")
```
Spatial Availability $V_i$ at resolution Zona 208
]

???
- a need for multi-modal analysis to assess impacts of this geographic. The issue of MUAP.
---
class: center
### Multi-modal home-to-work accessibility in Madrid: 2018 travel survey
```{r, echo=FALSE, out.width='45%',fig.align='center'}
knitr::include_graphics("Presentations/SA_im_V_zn1259_plot.png")
```
Spatial Availability $V_i^m$ at resolution Zona 1259

???
- With spatial availability you can compare between modes -- however still the issue of MUAP.

---
class: center
### Multi-modal home-to-work accessibility in Madrid: 2018 travel survey
```{r, echo=FALSE, out.width='45%',fig.align='center'}
knitr::include_graphics("Presentations/SA_im_vv_zn1259_plot.png")
```
Spatial Availability by $m$ population ( $v_i^m$ ) at resolution Zona 1259

???
- Can also benchmark between modes: the spatial avail per mode-using population per $i$ -- however still the issue of MUAP
---
### Spatial availability: a way forward in multimodal transport and land-use planning

- Data-driven transport accessibility policy scenario creation

- Modifiable unit areal problem (MUAP) - need for logically consistent accessibility measurement method
  - Calculated travel time tessellation

- Questions of equity: which neighbourhoods and people _within_ neighbourhoods are not meeting the mark? 
  - Who will be most impacted?

???
- policy scenario creation
- MUAP exists, seen in Madrid transport data, the need for simulated calculation and consistent aggregation of accessibility (through proportional allocation)
- Quantitatively, this method can be used to more interpetble answer these questions of equity. Socio-economic impact, GHG emissions, across sub-sets of population (multi-modal users included).

---
count: false
class: center, middle, inverse

@@

>>>>

---
title: "Quantifying multi-modal competition using spatial availability"
subtitle: "NECTAR - Sustainable Neighbourhoods"
author: "Anastasia Soukhov, J. Tarriño Ortiz, J. A. Soria-Lara, A. Páez"
date: "21-06-2023"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      slideNumberFormat: "%current% / 17" 
      highlightStyle: github
      highlightLines: true
      ratio: '16:9'
      countIncrementalSlides: false
      beforeInit: "macros.js"
---
```{r xaringan-themer, include=FALSE, warning=FALSE}
#this themer package allows us to change the .css themes through this chunk. Easier than changing the .css directly.
library(xaringanthemer)
style_mono_accent(
  base_color = "#7A003C", #maroon
  link_color = "#5E6A71", #grey
  text_bold_color = "#FDBF57", #gold
  text_slide_number_color = "#FDBF57", #gold
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "300", "300i"),
  code_font_google   = google_font("Fira Mono")
)
```

```{r load-packages, include=FALSE, cache=FALSE}
library(dplyr)
library(sf)
library(ggplot2)
library(tidyverse)
library(units)
library(patchwork)
library(ggspatial)
library(ggforce)
library(geomtextpath)
library(knitr)
library(kableExtra)
library("ggpmisc")
options(scipen = 999)
```

@@

#write the dataframe long.
od_tt_m <- data.frame(i = c("A", "A", "A", "A", "A", "A",
                            "B", "B", "B", "B", "B", "B",
                            "C", "C", "C", "C", "C", "C"), # Three origins
                    j = c("1", "2", "3", "1", "2", "3", 
                          "1", "2", "3", "1", "2", "3", 
                          "1", "2", "3", "1", "2", "3"), # Three destinations
                    m = c("pt", "pt", "pt", "car", "car", "car",
                          "pt", "pt", "pt", "car", "car", "car",
                          "pt", "pt", "pt", "car", "car", "car"),
                    tt = c(15, 30, 100, 10, 25, 80, #travel time, z-mode (original) AND THEN x-mode (the faster mode)
                           30, 15, 100, 25,  10,  80,
                           100, 100, 15, 80, 80, 10), 
                    pop = c(16666.66666, 16666.66666, 16666.66666, 33333.33333, 33333.33333, 33333.33333,
                            60000, 60000,60000, 90000, 90000, 90000,
                            3000, 3000, 3000, 7000,7000, 7000),
                    opp = c(100000, 100000, 10000, 100000, 100000, 10000,
                            100000, 100000, 10000, 100000, 100000, 10000,
                            100000, 100000, 10000, 100000, 100000, 10000)) 
```

@@

od_table <- data.frame(Center = c("A", "B", "C", "1", "2", "3"),
  Size = c("50,000 pop.", "150,000 pop.", "10,000 pop.", 
           "100,000 jobs", "100,000 jobs", "10,000 jobs"))

@@

```{r, create-figure-with-toy-example, echo=FALSE, eval=FALSE}
toy_example_plot <- ggplot() + 
  # Plot centers
  geom_circle(data = centers,
             aes(x0 = x, 
                 y0 = y,
                 r = radius),
             color = "grey40") +
  annotate(geom = "label", 
           x=c(1.6, 1.7, 7.5),
           y=c(9.1, 0.9, 5.2), 
           size=4, 
           label.size=NA) +
  # Connect origins and destinations with curves
  geom_curve(data = od_lines,
               aes(x = x,
                   y = y,
                   xend = xend,
                   yend = yend,
                   linetype = group,
                   group = group),
             curvature = 0.20,
             linewidth = 1,
             color = "black") +
  # Plot origins and destinations
  geom_point(data = od,
            aes(size = size,
                x = x, 
                y = y, 
                color = type,
                shape = type)) +
  scale_size(range = c(6, 16)) +
  # Label origins and destinations
  geom_text(data = od, 
            aes(x, 
                y, 
                label=id), 
            size=4) +
  # Create table
  annotate(geom = 'table',
           x = 9,
           y = 9,
           label = list(od_table),
           size=4) +
  # Theme the plot
  coord_fixed() + 
  theme_void() + guides(shape = "none", size = "none", color = "none", linetype ="none")

toy_example_plot
ggsave("Presentations/toy-example1.png",dpi=600)
```

## Accessibility competition between modes

**Motivation**: interpreting accessibility measures (location-based _potential interaction_) for policy interventions

```{r, toy-example0, out.width='40%', fig.align='center', echo=FALSE}
knitr::include_graphics("Presentations/access-demo.png")
```

---
## Accessibility competition between modes

**Motivation**: interpreting accessibility measures (location-based _potential interaction_) for policy interventions

```{r, toy-example00, out.width='40%', fig.align='center', echo=FALSE}
knitr::include_graphics("Presentations/access-demo.png")
```

### Overview

**1.** Types of current accessibility measures and their limitations

**2.** Multi-modal **spatial availability**

**3.** Planning accessibility policy interventions considering multi-modal competition

---
class: center, middle
## Classification of multi-modal accessibility measures

##### Hansen-type accessibility _(Hansen, 1959)_
_No competition_ and _unconstrained_: $S_i^m = \sum_j O_j f^m(c_{ij}^m)$

##### Shen-type accessibility _(Shen, 1998)_ and _(Luo and Wang, 2003)_
**Competitive** and _unconstrained_: $a_i^m = \sum_j \frac{O_jf^m(c_{ij}^m)}{\sum_m D_j^m}$

##### Spatial availability _(Soukhov et al., 2023)_
**Competitive** and **constrained**: $V_i^m = \sum_j O_j F^{t,m}_{ij}$

<!-- for this slide-->

Accessibility: a way to quantify the access that a spatial unit has, a product of land-use and the population's means of transportation. It measures the amount of _potential interaction_ a population has with opportunities _(Hansen, 1959)_

What do all these types of measures have in common? How they are a weighted sum of opportunities that can be interacted with. What differs between them is how they weight the opportunities.

### Simulated example: uncompetitive and unconstrained

@@

```{r, toy-example1, out.width='120%', echo=FALSE}
knitr::include_graphics("Presentations/toy-example1.png")
```

@@

##### Hansen-type accessibility
$S_i^m = \sum_jO_jf^m(c_{ij}^m)$

Impedance function: $f(c_{ij})^{\text{car}} = f(c_{ij})^{\text{pt}} =exp(-0.1 \cdot c_{ij}^m)$;

@@

S <- od_tt_m |>
  group_by(i,m) |>
  summarize(S = sum(opp * f),
            .groups = "drop")

add_data <- data.frame(i = "TOTAL", m = "" , S = S$S%>%sum())
S <- bind_rows(S, add_data)

S %>% kable(format = "html",
        booktabs = TRUE,
        digits = 1)
```
]

???
- NOTE: the sum of all that is what it means to be unconstrained. Also, A and B are equal, that is what it means to be uncompetitive.

- explain the map; blue and orange dots are origins and employment centers respectively. Travel time within each center is 15 mins, travel time between sub and urban is 30 mins and travel time to the satellite town from sub / urban is 100 mins.

@@

### Simulated example: competitive and unconstrained

.pull-left[
```{r, out.width='120%',echo=FALSE}
knitr::include_graphics("Presentations/toy-example1.png")
```
]

@@

##### Shen-type accessibility
$a_i^m = \sum_j \frac{O_j f^m(c_{ij}^m)}{\sum_m \sum_i P_{i}^m f^m(c_{ij}^m)}$

@@

a_m <- od_tt_m %>%
  left_join(LOS_j |> 
              dplyr::select(j, LOS_j, opp, D_j),
            by = c("j")) %>%
  group_by(i,m) |>
  summarize(a_i_m = sum(LOS_j * f),
            pop = mean(pop),
            .groups = "drop") |>
  rename("a" = "a_i_m")

```
```{r, echo=FALSE, message=FALSE}
knitr::kable(a_m %>% dplyr::select(-c("pop")),
             format = "html",
             booktabs = TRUE,
             digits = 2)
```
]

???
-  ## NOTE: this is competitive for S. you have high access with less demand, A and B are different. It is still unconstrained when you try to retrieve the number of opportunities -- it is the totla population and not the job-seeking population 
---
### Simulated example: competitive and unconstrained

.pull-left[
```{r, out.width='120%',echo=FALSE}
knitr::include_graphics("Presentations/toy-example1.png")
```

@@

##### Shen-type accessibility
$a_i = \sum_m \biggl(\frac{P_i^m}{P_i}\biggr) a_i^m$
```{r, echo=FALSE, message=FALSE}
a <- a_m %>%
  group_by(i) |>
  summarize(a = sum(pop*a)/sum(pop),
            .groups = "drop")

knitr::kable(a,
             format = "html",
             booktabs = TRUE,
             digits = 2)

# weighted average-- is based just on the population overall not the opportunity-skeeing populationg (i.e., not weighted by travel behaviour impedance function)
```

@@

---
### Spatial availability: competitive _and_ constrained

@@

- The **proportional allocation factor** combines two aspects of the problem, population and cost:

$F^{t,m}_{ij} = \frac{F^{pm}_{i} \cdot F^{cm}_{ij}}{\sum_{m} \sum_{i} F^{pm}_{i} \cdot F^{cm}_{ij}}$

- Where $F^{p,m}_{i}$ is **population allocation** factor:

$F^{p,m}_{i} = \frac{P^m_{i}}{\sum_{m}\sum_{i} P^m_{i}}$

- And the $F^{c,m}_{ij}$ is a **cost allocation** factor:

$F^{c,m}_{ij} = \frac{f^m(c_{ij}^m)}{\sum_{m} \sum_{i} f^m(c_{ij}^m)}$
]

@@

##### Proportional allocation mechanism

- $V^m_{i}$ is always singly-constrained: $\sum_{i} V_{i} = \sum_{m}\sum_{i} V_{i}^m = \sum_{j} O_j$

- So, one can calculate spatial availability....
  - **per $i$**: ( $V_{i} = \sum_{m} V^m_{i}$ )
  - **per mode**: ( $v^{m} = \sum_{m} \sum_{i} V^m_{i}$ )
  - **per capita in $i$**: ( $v_{i} = \sum_{m} \frac{V_i^m}{P_{i^m}}$ )
  - Etc. etc.,

@@

-supply of population and demand for opportunities are both travel-cost and population adjusted

- why singly-constrained? the sum of all accessibility values for i equals all the opportunities for i to j (like in a singly constrained gravity model - it is attraction constrained)

- Can calculate spatial availability per $i$, or per capita in $i$, or per $m$. Per subset of population -- anything! This is also something that you couldn't just do with the Shen-type measure because it is unconstrained.

---
### Simulated example: competitive _and_ constrained

```{r, out.width='120%',echo=FALSE}
knitr::include_graphics("Presentations/toy-example1.png")
```
]

@@

##### Spatial availability per origin and mode
$V^m_{i} = \sum_{j} O_j\ F^{t,m}_{ij}$

@@

V_m <- od_tt_m %>%
  group_by(i,m) %>%
  summarize(V = sum(V_ij),
            v = V/first(pop))

add_data <- data.frame(i = "TOTAL", m = "", V = V_m$V%>%sum(), v = NA)
V_m <- bind_rows(V_m, add_data)

```
```{r, echo=FALSE, message=FALSE}
knitr::kable(V_m %>% dplyr::select(-v),
             format = "html",
             booktabs = TRUE,
             digits = 2)
```

@@

### Simulated example: competitive _and_ constrained

.pull-left[
```{r,out.width='120%', echo=FALSE}
knitr::include_graphics("Presentations/toy-example1.png")
```
]

@@

##### Spatial availability per origin, per mode
$V_{i} = \sum_{m} V^m_{i}$

@@

V <- V %>% left_join(a, by=c("i")) |> rename("v" = "a") |> transmute(i, V, v)

add_data <- data.frame(i = "TOTAL", V = V$V%>%sum(), v = NA)
V <- bind_rows(V, add_data)
```

```{r, echo=FALSE, message=FALSE}
knitr::kable(V,
             format = "html",
             digits = 2)
```

$V^m = \sum_iV_i^m$

```{r, echo=FALSE, message=FALSE}
V_m <- od_tt_m %>%
  group_by(m) %>%
  summarize(V = sum(V_ij)) |> 
  transmute(m, V, 
            v = c((30+100+7)/(50+150+10), (20+50+3)/(50+150+10)))

add_data <- data.frame(m = "TOTAL", V = V_m$V%>%sum())
V_m <- bind_rows(V_m, add_data)

knitr::kable(V_m,
             format = "html",
             booktabs = TRUE,
             digits = 2) 
```

@@

.pull-right[
```{r, echo=FALSE, out.width='50%', fig.align='center'}
knitr::include_graphics("Presentations/city-centre.png")
```
- November 2018 - LEZ Centro
- 2023 - LEZ within M-30
- 2024 - LEZ within Madrid Municipality 
]

???
A NEEED for multi-modal analysis methods that are interpretive and comparative -- proportional allocation achieves that!
---
class: center
### Multi-modal home-to-work accessibility in Madrid: 2018 travel survey
```{r, echo=FALSE, out.width='70%',fig.align='center'}
knitr::include_graphics("Presentations/im_populations_zn208_plot.png")
```
Worker population, by mode, at resolution Zona 208

---
class: center
### Multi-modal home-to-work accessibility in Madrid: 2018 travel survey
```{r, echo=FALSE, out.width='70%',fig.align='center'}
knitr::include_graphics("Presentations/Hansen_im_V_zn208_plot.png")
```
Uncompetitive and unconstrained accessibility $S_i^m$ at resolution Zona 208

???
1) totals don't add up - not constrained
2) uncompetive between modes

---
class: center
### Multi-modal home-to-work accessibility in Madrid: 2018 travel survey
```{r, echo=FALSE, out.width='70%',fig.align='center'}
knitr::include_graphics("Presentations/SA_im_V_zn208_plot.png")
```
Spatial Availability $V_i^m$ at resolution Zona 208

???
- SA is better for transit/walk for centro. Car mode is limited in the centro. 
- those who live in the periphery have high accessibility, those are the ones who have cars and use transit.

---
class: center
### Multi-modal home-to-work accessibility in Madrid: 2018 travel survey
```{r, echo=FALSE, out.width='70%', fig.align='center'}
knitr::include_graphics("Presentations/SA_i_V_zn208_plot.png")
```
Spatial Availability $V_i$ at resolution Zona 208

???
- center -- high competition, for all modes and those that travel from close/far of the city.

---
class: middle
### Spatial availability: planning multi-modal transportation policy

- **Scenario evaluation**: accessibility when...
  - LEZ boundaries change
  - Vehicle characteristic restrictions change
  - Travel behaviour changes
  - Spatial unit of analysis changes

- **Questions of equity**: which neighbourhoods, what people, what modes?
  - GHG emission reductions
  - Advantages to non-car users
  - Who are impacted disproportionally

???
- policy scenario creation

- Quantitatively, this method can be used to more interpetble answer these questions of equity. Socio-economic impact, GHG emissions, across sub-sets of population (multi-modal users included).

<!-- - MUAP exists, seen in Madrid transport data, the need for simulated calculation and consistent aggregation of accessibility (through proportional allocation) -->

@@

---
---
## Multi-modal accessibility measures

#### No competition and unconstrained
##### Hansen-type accessibility
